---
title: "STA222 FP"
output: html_document
date: "2023-11-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Preprocessing

```{r}
library(KMsurv)
library(survival)
library(ggplot2)
library(dplyr)
library(MASS)

data("cancer")
rotterdam1 <- rotterdam

#rotterdam1 <- rotterdam1[rotterdam1$er != 0 | rotterdam1$pgr != 0, ]


## Convert categorical variables to factors
rotterdam1$year <- factor(rotterdam1$year)
rotterdam1$meno <- factor(rotterdam1$meno, levels = c(0, 1), 
                          labels = c("premenopausal", "postmenopausal"))
rotterdam1$hormon <- factor(rotterdam1$hormon, levels = c(0, 1), 
                          labels = c("No", "Yes"))
rotterdam1$chemo <- factor(rotterdam1$chemo, levels = c(0, 1), 
                          labels = c("No", "Yes"))
rotterdam1$grade <- factor(rotterdam1$grade)

rotterdam1$age_cate <- cut(rotterdam1$age, breaks = c(-Inf, 40, 60, Inf), 
                           labels = c("20-40", "40-60", "above 60"))

rotterdam1$nodes_cate <- cut(rotterdam1$nodes, breaks = c(-Inf, 1, 4, Inf), 
                           labels = c("0-1", "2-4", "above 4"))
summary(rotterdam1$nodes_cate)

rotterdam1$er_cate <- cut(rotterdam1$er, breaks = c(-Inf, 20, 130, Inf), 
                           labels = c("0-20", "21-130", "above 130"))
summary(rotterdam1$er_cate)

rotterdam1$pgr_cate <- cut(rotterdam1$pgr, breaks = c(-Inf, 10, 120, Inf), 
                           labels = c("0-10", "11-120", "above 120"))
summary(rotterdam1$pgr_cate)

## New dataset
#set the final observation time, dtime = Time To Death Or Last follow-up
rotterdam2 <- tmerge(rotterdam1, rotterdam1, pid, tstop=dtime)
#set time until relapse
rotterdam2 <- tmerge(rotterdam2, rotterdam1, pid, newrecur=tdc(rtime))
#status at the end, either they dead or they were censored
status <- as.integer(with(rotterdam2, (tstop == dtime & death)))
#put together
rotterdam2 <- data.frame(rotterdam2,status)

# Build up the survial object
## deathtime-death
rotterdam1.surv <- with(rotterdam1,Surv(dtime,death))

rotterdam2.surv <- with(rotterdam2,Surv(time=tstart,time2=tstop, event=status,type="counting"))


```

```{r}
summary(rotterdam1$er)
summary(rotterdam1$pgr)
rotterdam1_ER0 <- rotterdam1[rotterdam1$er < 1, ]
rotterdam1_PGR0 <- rotterdam1[rotterdam1$pgr < 1, ]
rotterdam1_R0 <- rotterdam1[rotterdam1$er <= 1 & rotterdam1$pgr >= 0, ]

rotterdam1_R1 <- rotterdam1[rotterdam1$er != 0 | rotterdam1$pgr != 0, ]

```
# ############################################################################
## Survivial Analysis

### Plot the Kaplan-Meier curves for the hormonal treated and untreated patients and use survdiff to test for whether the curves are different.

```{r}
plot(survfit(rotterdam1.surv~hormon,data=rotterdam1),col=1:2,lwd=2)
title("Time to Death for No hormonal treatment and Hormonal treatment")
legend("topright",c("No hormonal treatment","Hormonal treatment"),col=1:2,lwd=2)

survdiff(formula = rotterdam1.surv ~ hormon, data = rotterdam1)

```

### Plot the cumulative hazards vs. time and the complimentary log-log survival vs. log time.

```{r}
plot(survfit(rotterdam1.surv~hormon,data=rotterdam1),col=1:2,lwd=2,fun="cumhaz")
title("Cumulative Hazards vs. Time")
legend("topleft",c("No hormonal treatment","Hormonal treatment"),col=1:2,lwd=2)

plot(survfit(rotterdam1.surv~hormon,data=rotterdam1), fun="cloglog", 
     xlab="Log(Time)", ylab="Log[-Log(Survival Probability)]", col=1:2, lwd=2)
title("Complimentary log-log Survival vs. log Time")
legend("topleft",c("No hormonal treatment","Hormonal treatment"),col=1:2,lwd=2)
```

### Hazard Ratios for hormon treatment

```{r}
KMcurves <- survfit(rotterdam1.surv~hormon,data=rotterdam1)
timevec <- 1:7500
#group No
sf1 <- stepfun(KMcurves[1]$time,c(1,KMcurves[1]$surv))
#group Yes
sf2 <- stepfun(KMcurves[2]$time,c(1,KMcurves[2]$surv))
cumhaz1 <- -log(sf1(timevec))
cumhaz2 <- -log(sf2(timevec))

plot(timevec,cumhaz1/cumhaz2,type="l",ylab="Hazard Ratio",xlab="Time",ylim=c(0,2))
legend("topright",c("No Hormonal treatment/Hormonal Treatment"), lty = 1)
title("Hazard Ratios for Hormonal Treatment")
```

### KM curves: check Partial Effects of Covariates

```{r}
plot(survfit(rotterdam1.surv~age_cate,data=rotterdam1),col=1:3,lwd=2)
title("Survival proportion for Age")
legend("topright",c("20-40", "40-60", "above 60"),col=1:3,lwd=2)

plot(survfit(rotterdam1.surv~meno,data=rotterdam1),col=1:2,lwd=2)
title("Survival proportion for menopausal status")
legend("topright",c("premenopausal", "postmenopausal"),col=1:2,lwd=2)

plot(survfit(rotterdam1.surv~size,data=rotterdam1),col=1:3,lwd=2)
title("Survival proportion for tumor size")
legend("topright",c("<=20","20-50",">50"),col=1:3,lwd=2)

plot(survfit(rotterdam1.surv~grade,data=rotterdam1),col=1:3,lwd=2)
title("Survival proportion for differentiation grade")
legend("topright",c("2","3","1"),col=1:3,lwd=2)

plot(survfit(rotterdam1.surv~nodes_cate,data=rotterdam1),col=1:3,lwd=2)
title("Survival proportion for number of positive lymph nodes")
legend("topright",c("0-1", "2-4", "above 4"),col=1:3,lwd=2)

plot(survfit(rotterdam1.surv~pgr_cate,data=rotterdam1),col=1:3,lwd=2)
title("Survival proportion for progesterone receptors (fmol/l)")
legend("topright",c("0-20", "21-130", "above 130"),col=1:3,lwd=2)

plot(survfit(rotterdam1.surv~er_cate,data=rotterdam1),col=1:3,lwd=2)
title("Survival proportion for estrogen receptors (fmol/l)")
legend("topright",c("0-10", "11-120", "above 120"),col=1:3,lwd=2)

plot(survfit(rotterdam1.surv~chemo,data=rotterdam1),col=1:2,lwd=2)
title("Survival proportion for chemotherapy")
legend("topright",c("No chemotherapy", "chemotherapy"),col=1:2,lwd=2)
```
# ############################################################################


## Model Building

```{r}
# Full model
cox_full <- coxph(rotterdam1.surv ~ hormon*chemo+year+meno+size+pgr+er+nodes+grade+age, data = rotterdam1)
summary(cox_full)
AIC(cox_full)
## year, meno, er and hoemon*chemo are not significant

## Drop variables interpretations
## year: The cancer is only based on your gender and age, not highly related to the year of surgery.
## meno: Since meno is highly correlated with the age, we will also drop this variables
## Node: Due to the unbalance distribution (Right Skewed)

## Stratification Interpretation
### Because size, pgr and er(after evenly cut) have intersection for the survival curves ,we do the stratification.

```

## Time-Independent Cox Model

```{r}
### After stepwise selection, the final time-dependent model is showed below.
cox1 <- coxph(rotterdam1.surv ~ hormon*chemo+strata(size)+strata(pgr_cate)+strata(er_cate)+grade+age, data = rotterdam1)
summary(cox1)
# All predictors are significant even for the interaction of chemo and meno.

AIC(cox1) 
# AIC= 11076.21

cox1.zph <- cox.zph(cox1)
cox1.zph
#From the score test, there is no significant evidence to show that non-proportionality for the predictors hormon, chem and their interaction, grade, age. All of their p-values are greater than 0.05.

rotterdam1$hormon*rotterdam1$chemo
```

## Time-Dependent Cox Model

```{r}
cox2 <- coxph(rotterdam2.surv ~ hormon*chemo+strata(size)+strata(pgr_cate)+strata(er_cate)+grade+age+newrecur, data = rotterdam2)
summary(cox2)
# The grade and the interaction of chemo and hormon are not significant for the time-dependent model.

AIC(cox2)
# AIC = 8700.053


cox2.zph <- cox.zph(cox2)
cox2.zph
#From the score test, there is no significant evidence to show that non-proportionality for the predictors hormon, chem and their interaction, grade, age. All of their p-values are greater than 0.05. However, recurrence might have an non-proportionality due to its lower p-value.

```

### Model Checking

```{r}
# For time-independent cox model
plot(cox1.zph[1], main = "Schoenfeld Residuals for Hormonal")
plot(cox1.zph[2], main = "Schoenfeld Residuals for Chemotherapy")
plot(cox1.zph[3], main = "Schoenfeld Residuals for Grade")
plot(cox1.zph[4], main = "Schoenfeld Residuals for Age")
plot(cox1.zph[5], main = "Schoenfeld Residuals for Interaction Between Hormonal Treatment and Chemotherapy")

```

```{r}
rotterdam.mart <- residuals(cox1, type="martingale")
rotterdam.cs <- rotterdam1$death-rotterdam.mart

surv.csr <- survfit(Surv(rotterdam.cs, rotterdam1$death)~1,type="fleming-harrington")
plot(surv.csr,fun="cumhaz")
abline(0,1)
title("Cumulative Hazard of Cox-Snell Residuals")

```


```{r}
#fit residuals: martingale, deviance, and df beta
rotterdam.mart <- residuals(cox1, type="martingale")
rotterdam.dev  <- residuals(cox1, type="deviance") 
rotterdam.dfb  <- residuals(cox1, type="dfbeta")
rotterdam.preds <- predict(cox1)

#the martingale residuals vs. the linear predictor
plot(rotterdam.preds, rotterdam.mart, xlab="Linear Predictor", 
     ylab="Martingale Residual", ylim = c(-3, 1.5), pch = 19, cex = 0.5)
title("Martingale Residuals vs. Linear Predictor") 
text(rotterdam.preds, rotterdam.mart+0.2, labels = rownames(rotterdam2))

#the deviance residuals vs. the linear predictor
plot(rotterdam.preds, rotterdam.dev, xlab="Linear Predictor", 
     ylab="Deviance Residual", ylim = c(-3,4), pch = 19, cex = 0.5)
title("Deviance Residuals vs. Linear Predictor") 
text(rotterdam.preds, rotterdam.dev+0.2, labels = rownames(rotterdam2))

```