---
title: "STA222 FP"
output: html_document
date: "2023-11-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Preprocessing

```{r}
library(KMsurv)
library(survival)
library(ggplot2)
library(dplyr)
library(MASS)

data("cancer")
rotterdam1 <- rotterdam

## Convert categorical variables to factors
rotterdam1$year <- factor(rotterdam1$year)
rotterdam1$meno <- factor(rotterdam1$meno, levels = c(0, 1), 
                          labels = c("premenopausal", "postmenopausal"))
rotterdam1$hormon <- factor(rotterdam1$hormon, levels = c(0, 1), 
                          labels = c("No", "Yes"))
rotterdam1$chemo <- factor(rotterdam1$chemo, levels = c(0, 1), 
                          labels = c("No", "Yes"))
rotterdam1$grade <- factor(rotterdam1$grade)


rotterdam1$nodes_cate <- cut(rotterdam1$nodes, breaks = c(-Inf, 1, 4, Inf), 
                           labels = c("0-1", "2-4", "above 4"))
summary(rotterdam1$nodes_cate)

rotterdam1$er_cate <- cut(rotterdam1$er, breaks = c(-Inf, 20, 130, Inf), 
                           labels = c("0-20", "21-130", "above 130"))
summary(rotterdam1$er_cate)

rotterdam1$pgr_cate <- cut(rotterdam1$pgr, breaks = c(-Inf, 10, 120, Inf), 
                           labels = c("0-10", "11-120", "above 120"))
summary(rotterdam1$pgr_cate)

## New dataset
#set the final observation time, dtime = Time To Death Or Last follow-up
rotterdam2 <- tmerge(rotterdam1, rotterdam1, pid, tstop=dtime)
#set time until relapse
rotterdam2 <- tmerge(rotterdam2, rotterdam1, pid, newrecur=tdc(rtime))
#status at the end, either they dead or they were not censored
status <- as.integer(with(rotterdam2, (tstop == dtime & death)))
#put together
rotterdam2 <- data.frame(rotterdam2,status)

# Build up the survial object
## deathtime-death
rotterdam1.surv <- with(rotterdam1,Surv(dtime,death))

rotterdam2.surv <- with(rotterdam2,Surv(time=tstart,time2=tstop, event=status,type="counting"))

summary(rotterdam1)
str(rotterdam1)

summary(rotterdam2)
str(rotterdam2)
```

## Survivial Analysis

```{r}
plot(survfit(rotterdam1.surv~nodes_cate,data=rotterdam1),col=1:3,lwd=2)
title("Survival proportion for Nodes")
legend("topright",c("0-1", "2-4", "above 4"),col=1:3,lwd=2)

plot(survfit(rotterdam1.surv~pgr_cate,data=rotterdam1),col=1:3,lwd=2)
title("Survival proportion for pgr")
legend("topright",c("0-20", "21-130", "above 130"),col=1:3,lwd=2)

plot(survfit(rotterdam1.surv~er_cate,data=rotterdam1),col=1:3,lwd=2)
title("Survival proportion for er")
legend("topright",c("0-10", "11-120", "above 120"),col=1:3,lwd=2)

plot(survfit(rotterdam1.surv~size,data=rotterdam1),col=1:3,lwd=2)
title("Survival proportion for Size")
legend("topright",c("<=20","20-50",">50"),col=1:3,lwd=2)


```

## Model Building

```{r}
# Full model
cox_full <- coxph(rotterdam1.surv ~ hormon*chemo+year+meno+size+pgr+er+nodes+grade+age, data = rotterdam1)
summary(cox_full)
AIC(cox_full)
## year, meno, er and hoemon*chemo are not significant

## Drop variables interpretations
## year: The cancer is only based on your gender and age, not highly related to the year of surgery.
## meno: Since meno is highly correlated with the age, we will also drop this variables
## Node: Due to the unbalance distribution (Right Skewed)

## Stratification Interpretation
### Because size, pgr and er(after evenly cut) have intersection for the survival curves ,we do the stratification.

```

## Time-Independent Cox Model

```{r}
### After stepwise selection, the final time-dependent model is showed below.
cox1 <- coxph(rotterdam1.surv ~ hormon*chemo+strata(size)+strata(pgr_cate)+strata(er_cate)+grade+age, data = rotterdam1)
summary(cox1)
# All predictors are significant even for the interaction of chemo and meno.

AIC(cox1) 
# AIC= 11076.21

cox1.zph <- cox.zph(cox1)
cox1.zph
#From the score test, there is no significant evidence to show that non-proportionality for the predictors hormon, chem and their interaction, grade, age. All of their p-values are greater than 0.05.

plot(cox1.zph)


#fit without age
cox2 <- coxph(rotterdam1.surv ~ hormon*chemo+strata(size)+strata(pgr_cate)+strata(er_cate)+grade, data = rotterdam1) #find residuals
mres <- residuals(cox2, type = "martingale")
#plot score vs residuals
plot(rotterdam1$age,mres,xlab="Age",ylab="Martingale Residuals") 
lines(lowess(rotterdam1$age,mres))
title("Martingale Residuals vs. Age")
```

## Time-Dependent Cox Model

```{r}
cox2 <- coxph(rotterdam2.surv ~ hormon*chemo+strata(size)+strata(pgr_cate)+strata(er_cate)+grade+age+newrecur, data = rotterdam2)
summary(cox2)
# The grade and the interaction of chemo and hormon are not significant for the time-dependent model.

AIC(cox2)
# AIC = 8700.053


cox2.zph <- cox.zph(cox2)
cox2.zph
#From the score test, there is no significant evidence to show that non-proportionality for the predictors hormon, chem and their interaction, grade, age. All of their p-values are greater than 0.05. However, recurrence might have an non-proportionality due to its lower p-value.


# Time-Indep
#fit without age
cox3 <- coxph(rotterdam1.surv ~ hormon*chemo+strata(size)+strata(pgr_cate)+strata(er_cate)+grade, data = rotterdam1) #find residuals
mres <- residuals(cox3, type = "martingale")
#plot score vs residuals
plot(rotterdam1$age,mres,xlab="Age",ylab="Martingale Residuals") 
lines(lowess(rotterdam1$age,mres))
title("Martingale Residuals vs. Age")

# Time-Dep
#fit without age
cox4 <- coxph(rotterdam2.surv ~ hormon*chemo+strata(size)+strata(pgr_cate)+strata(er_cate)+grade+newrecur, data = rotterdam2) #find residuals
mres3 <- residuals(cox4, type = "martingale")
#plot score vs residuals
plot(rotterdam2$age, mres3, xlab="Age", ylab="Martingale Residuals") 
lines(lowess(rotterdam2$age,mres3))
title("Martingale Residuals vs. Age")
```


### Model Checking

```{r}
# For time-independent cox model
plot(cox1.zph[1], main = "Schoenfeld Residuals for Hormonal Treatment")
plot(cox1.zph[2], main = "Schoenfeld Residuals for Chemotherapy")
plot(cox1.zph[3], main = "Schoenfeld Residuals for Grade")
plot(cox1.zph[4], main = "Schoenfeld Residuals for Age")
plot(cox1.zph[5], main = "Schoenfeld Residuals for Treatment Interaction")

```

```{r}
rotterdam.mart <- residuals(cox1, type="martingale")
rotterdam.cs <- rotterdam1$death-rotterdam.mart

surv.csr <- survfit(Surv(rotterdam.cs, rotterdam1$death)~1,type="fleming-harrington")
plot(surv.csr,fun="cumhaz")
abline(0,1)
title("Cumulative Hazard of Cox-Snell Residuals")


rotterdam.mart2 <- residuals(cox2, type="martingale")
rotterdam.cs2 <- rotterdam2$death-rotterdam.mart2

surv.csr2 <- survfit(Surv(rotterdam.cs2, rotterdam2$death)~1,type="fleming-harrington")
plot(surv.csr2,fun="cumhaz")
abline(0,1)
title("Cumulative Hazard of Cox-Snell Residuals")

```


```{r}
#fit residuals: martingale, deviance, and df beta
rotterdam.mart <- residuals(cox1, type="martingale")
rotterdam.dev  <- residuals(cox1, type="deviance") 
rotterdam.dfb  <- residuals(cox1, type="dfbeta")
rotterdam.preds <- predict(cox1)

#the martingale residuals vs. the linear predictor
plot(rotterdam.preds, rotterdam.mart, xlab="Linear Predictor", 
     ylab="Martingale Residual", ylim = c(-3, 1.5), pch = 16, cex = 0.5, col = "orange")
#title("Martingale Residuals vs. Linear Predictor") 
text(rotterdam.preds, rotterdam.mart+0.2, labels = rownames(rotterdam1), cex = 0.7)

#the deviance residuals vs. the linear predictor
plot(rotterdam.preds, rotterdam.dev, xlab="Linear Predictor", 
     ylab="Deviance Residual", ylim = c(-2.5,3.5), pch = 16, cex = 0.5,  col = "orange")
#title("Deviance Residuals vs. Linear Predictor") 
text(rotterdam.preds, rotterdam.dev+0.2, labels = rownames(rotterdam1),cex = 0.7)


plot(rotterdam.dfb[,1],xlab="Observation Number", ylab="dfbeta for Hormonal Treatment",
ylim=c(-0.01,0.01), pch = 16, cex = 0.5,  col = "orange")
#title("dfbeta Values by Observation Number for Hormonal Treatment") 
text(rotterdam.dfb[,1]+0.002, labels = rownames(rotterdam1),cex = 0.7)

plot(rotterdam.dfb[,2],xlab="Observation Number", ylab="dfbeta for Chemotherapy",
ylim=c(-0.01,0.01), pch = 16, cex = 0.5, col = "orange")
#title("dfbeta Values by Observation Number for Chemotherapy") 
text(rotterdam.dfb[,2]+0.002, labels = rownames(rotterdam1),cex = 0.7)

plot(rotterdam.dfb[,3],xlab="Observation Number", ylab="dfbeta for Differentiation Grade",
ylim=c(-0.005,0.01), pch = 19, cex = 0.5, col = "orange")
#title("dfbeta Values by Observation Number for Differentiation Grade") 
text(rotterdam.dfb[,3]+0.002, labels = rownames(rotterdam1), cex = 0.7)

plot(rotterdam.dfb[,4],xlab="Observation Number", ylab="dfbeta for Age",
ylim=c(0.0017,0.0022), pch = 19, cex = 0.5, col = "orange")
#title("dfbeta Values by Observation Number for Age") 
text(rotterdam.dfb[,4]+0.002, labels = rownames(rotterdam1),cex = 0.7)

plot(rotterdam.dfb[,5],xlab="Observation Number", ylab="dfbeta for Treatment Interaction ",
ylim=c(-0.1,0.12), pch = 19, cex = 0.5, col = "orange")
#title("dfbeta Values by Observation Number for Treatment Interaction") 
text(rotterdam.dfb[,5]+0.002, labels = rownames(rotterdam1),cex = 0.7)

```

### Observations to Examine by Residuals and Influence:

- **Martingale Residuals:** 2277, 2358
- **Deviance Residuals:** 2818
- **Hormonal Treatment Influence:** 2146
- **Chemotherapy Influence:** 2571, 1482, 2277
- **Grade Influence:** 476, 2046
- **Age Influence:** 2872, 2941
- **Interaction Influence:** 2085, 1091, 1820
The most important observations to examine seem to be 2277

```{r}
unusuals <- c(2277) 
rotterdam1[unusuals, ]
```




